---
title: "Properties of Bootstrap Confidence Intervals"
author: "Math 437 Spring 2023"
date: "2/13/2023"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Let's investigate the different properties of three bootstrap confidence intervals (percentile, basic, and normal-theory) and compare them to the t confidence interval. We will focus on bootstrap estimation of the sample mean so that we don't have to worry about iterated bootstrap.

```{r relevant packages}
library(dplyr)
library(ggplot2) # optional for Part 5
```

## Constructing the Bootstrap Confidence Intervals

Rather than using the built-in functions in the `boot` package, let's make sure we understand what those functions are doing by building our own functions.

First we need to do the bootstrap resampling. This function will do it once filled in properly:

```{r bootstrap resample function}
bootstrap_resample <- function(data_vector, B, summary_fn = mean,seed=100, ...){
  # data_vector: a vector of data
  # B: the number of bootstrap resamples
  # summary_fn: the name of a function to apply to the resampled data
  # ...: any additional arguments to summary_fn
  
  boot_samples <- matrix(0, nrow = length(data_vector), ncol = B)
  
  n <- length(data_vector)
  # We need to add some code here!
  set.seed(seed)
  
  for (i in 1:B) {
    boot_samples[,i] <- sample(data_vector, size = n, replace = TRUE)

  }
  
  boot_stat <- apply(boot_samples, 2, summary_fn, ...)
  # Seriously, do not name any arguments x or FUN when using apply within a function
  # Advice from someone who spent 30 minutes debugging a sample solution
  
  return(boot_stat)
}
```

Then we can create our confidence interval from the bootstrap-resampled estimates. You need to fill in the code in the `if/else` section to compute the bootstrap confidence interval the appropriate way.

```{r bootstrap CI function}
bootstrap_ci <- function(data_vector, method, B = 1000, seed = 100, C = 0.95, summary_fn = mean, ...){
  # data_vector: a vector of data
  # method: the CI method
  # B: the number of bootstrap resamples
  # seed: the seed to use
  # C: the confidence level as a decimal
  # summary_fn: the name of a function to apply to the resampled data
  # ...: any additional arguments to summary_fn


  
  obs_stat <- do.call(summary_fn, args = list(data_vector, ...))
  # do.call allows you to call a function without having to hard-code what that function is
  # the args argument is a list of arguments to the function
  # so this will find the observed value of the statistic given the original data vector
  
  bootstrap_values <- bootstrap_resample(data_vector, B, summary_fn, ...)
  
  alpha <- 1 - C
  if (method == "percentile"){
      # write code to get the percentile confidence level out of the returned bootstrap_values and store in a length 2 vector boot_ci
    boot_ci <- quantile(bootstrap_values, probs = c(alpha/2,1-alpha/2))
  } else if (method == "basic"){
    # write code to get the "basic" confidence level and store in a length 2 vector boot_ci
    boot_ci <- 2*obs_stat - quantile(bootstrap_values, probs = c(1-alpha/2 , alpha /2))
  } else if (method == "normal"){
    # write code to get the normal-theory confidence interval and store in a length 2 vector boot_ci
    center <- 2 * obs_stat - mean(bootstrap_values)
    crit_value <- qnorm(alpha/2)
    se_boot <- sd(bootstrap_values)
    boot_ci <- center + c(1, -1)*crit_value*se_boot
    # make sure to use the adjustments in the course notes, e.g., don't use 1/sqrt(n) as your standard deviation of sample means
  }
  
  return(boot_ci)
}
```

## Simulation Part 1 - Assumptions Met, Large Samples

Simulate 1000 sets of 100 random numbers from a standard normal distribution and store them in a 1000 x 100 matrix `sim_data1`. Note that each row of `sim_data1` represents a sample of size 100 under conditions where we *know* the assumptions of a t confidence interval are met.

```{r sim1}
set.seed(437)
# need to do the simulation now!
sim_data1 <- matrix(rnorm(1000*100), nrow = 1000, ncol = 100)
```

Let's confirm that we get somewhere around 95% coverage with our t confidence intervals:

```{r t-CI1}
# You should be able to just run this code chunk without any fixes
pop_mean1 <- 0
ci_t <- apply(sim_data1, 1, function(x) t.test(x)$conf.int)
ci_t_df <- as.data.frame(t(ci_t))
names(ci_t_df) <- c("lower", "upper")
ci_t_df %>% mutate(covered = lower <= pop_mean1 & upper >= pop_mean1) %>%
  summarize(coverage_probability = mean(covered))
```

Now, let's see how our bootstrap CI functions compare.

```{r boot percentile CI 1}
# You should be able to just run this code chunk without any fixes
ci_perc <- apply(sim_data1, 1, bootstrap_ci, method = "percentile", B = 1000, summary_fn = mean, na.rm = TRUE)
# don't need any ... arguments here, but illustrating the idea of the ...
# notice that bootstrap_ci has default seed = 100 and C = 0.95 arguments
ci_perc_df <- as.data.frame(t(ci_perc))
names(ci_perc_df) <- c("lower", "upper")
ci_perc_df %>% mutate(covered = lower <= pop_mean1 & upper >= pop_mean1) %>%
  summarize(coverage_probability = mean(covered))
```

Copy and modify the chunk above for the basic and normal-theory intervals.

```{r boot basic CI 1}
# You should be able to just run this code chunk without any fixes
ci_basic <- apply(sim_data1, 1, bootstrap_ci, method = "basic", B = 1000, summary_fn = mean, na.rm = TRUE)
# don't need any ... arguments here, but illustrating the idea of the ...
# notice that bootstrap_ci has default seed = 100 and C = 0.95 arguments
ci_basic_df <- as.data.frame(t(ci_basic))
names(ci_basic_df) <- c("lower", "upper")
ci_basic_df %>% mutate(covered = lower <= pop_mean1 & upper >= pop_mean1) %>%
  summarize(coverage_probability = mean(covered))
```

```{r boot normal CI 1}
# You should be able to just run this code chunk without any fixes
ci_norm <- apply(sim_data1, 1, bootstrap_ci, method = "normal", B = 1000, summary_fn = mean, na.rm = TRUE)
# don't need any ... arguments here, but illustrating the idea of the ...
# notice that bootstrap_ci has default seed = 100 and C = 0.95 arguments
ci_norm_df <- as.data.frame(t(ci_norm))
names(ci_norm_df) <- c("lower", "upper")
ci_norm_df %>% mutate(covered = lower <= pop_mean1 & upper >= pop_mean1) %>%
  summarize(coverage_probability = mean(covered))
```


What proportion of "95% confidence intervals" actually contained the true parameter value? Did any methods perform noticeably better/worse?

## Simulation Part 2 - Assumptions Met, Small Samples

Simulate 1000 sets of 10 random numbers from a standard normal distribution and store them in a 1000 x 100 matrix `sim_data2`. Note that `sim_data2` represents 1000 samples of size 10 under conditions where we *know* the assumptions of a t confidence interval are met.

Copy and modify your code chunks from Simulation Part 1 to produce 95% confidence intervals using `sim_data2`. What proportion of "95% confidence intervals" actually contained the true parameter value? Did any methods perform noticeably better/worse?

```{r sim 2}
set.seed(437)
# need to do the simulation now!
sim_data2 <- matrix(rnorm(1000*10), nrow = 1000, ncol = 10)
```

```{r t-CI2}
# You should be able to just run this code chunk without any fixes
pop_mean2 <- 0
ci_t2 <- apply(sim_data2, 1, function(x) t.test(x)$conf.int)
ci_t_df2 <- as.data.frame(t(ci_t2))
names(ci_t_df2) <- c("lower", "upper")
ci_t_df2 %>% mutate(covered = lower <= pop_mean2 & upper >= pop_mean2) %>%
  summarize(coverage_probability = mean(covered))
```

```{r boot percentile CI 2}
# You should be able to just run this code chunk without any fixes
ci_perc2 <- apply(sim_data2, 1, bootstrap_ci, method = "percentile", B = 1000, summary_fn = mean, na.rm = TRUE)
# don't need any ... arguments here, but illustrating the idea of the ...
# notice that bootstrap_ci has default seed = 100 and C = 0.95 arguments
ci_perc_df2 <- as.data.frame(t(ci_perc2))
names(ci_perc_df2) <- c("lower", "upper")
ci_perc_df2 %>% mutate(covered = lower <= pop_mean2 & upper >= pop_mean2) %>%
  summarize(coverage_probability = mean(covered))
```

```{r boot basic CI 2}
# You should be able to just run this code chunk without any fixes
ci_basic2 <- apply(sim_data2, 1, bootstrap_ci, method = "basic", B = 1000, summary_fn = mean, na.rm = TRUE)
# don't need any ... arguments here, but illustrating the idea of the ...
# notice that bootstrap_ci has default seed = 100 and C = 0.95 arguments
ci_basic_df2 <- as.data.frame(t(ci_basic2))
names(ci_basic_df2) <- c("lower", "upper")
ci_basic_df2 %>% mutate(covered = lower <= pop_mean2 & upper >= pop_mean2) %>%
  summarize(coverage_probability = mean(covered))
```

```{r boot normal CI 2}
# You should be able to just run this code chunk without any fixes
ci_norm2 <- apply(sim_data2, 1, bootstrap_ci, method = "normal", B = 1000, summary_fn = mean, na.rm = TRUE)
# don't need any ... arguments here, but illustrating the idea of the ...
# notice that bootstrap_ci has default seed = 100 and C = 0.95 arguments
ci_norm_df2 <- as.data.frame(t(ci_norm2))
names(ci_norm_df2) <- c("lower", "upper")
ci_norm_df2 %>% mutate(covered = lower <= pop_mean2 & upper >= pop_mean2) %>%
  summarize(coverage_probability = mean(covered))
```


## Simulation Part 3 - Assumptions Not Met, Large Samples

Simulate 1000 sets of 100 random numbers from a $Exp(1)$ distribution and store them in a 1000 x 100 matrix `sim_data3`. Note that `sim_data3` represents 1000 samples of size 100 under conditions where we *know* the assumptions of a t confidence interval are not met but we are hoping that the Central Limit Theorem can cover us.

Copy and modify your code chunks from Simulation Part 1 to produce 95% confidence intervals using `sim_data3`. What proportion of "95% confidence intervals" actually contained the true parameter value? Did any methods perform noticeably better/worse?

```{r sim 3}
set.seed(437)
# need to do the simulation now!
sim_data3 <- matrix(rexp(1000*100), nrow = 1000, ncol = 100)
```

```{r t-CI3}
# You should be able to just run this code chunk without any fixes
pop_mean3 <- 1
ci_t3 <- apply(sim_data3, 1, function(x) t.test(x)$conf.int)
ci_t_df3 <- as.data.frame(t(ci_t3))
names(ci_t_df3) <- c("lower", "upper")
ci_t_df3 %>% mutate(covered = lower <= pop_mean3 & upper >= pop_mean3) %>%
  summarize(coverage_probability = mean(covered))
```

```{r boot percentile CI 3}
# You should be able to just run this code chunk without any fixes
ci_perc3 <- apply(sim_data3, 1, bootstrap_ci, method = "percentile", B = 1000, summary_fn = mean, na.rm = TRUE)
# don't need any ... arguments here, but illustrating the idea of the ...
# notice that bootstrap_ci has default seed = 100 and C = 0.95 arguments
ci_perc_df3 <- as.data.frame(t(ci_perc3))
names(ci_perc_df3) <- c("lower", "upper")
ci_perc_df3 %>% mutate(covered = lower <= pop_mean3 & upper >= pop_mean3) %>%
  summarize(coverage_probability = mean(covered))
```

Copy and modify the chunk above for the basic and normal-theory intervals.

```{r boot basic CI 3}
# You should be able to just run this code chunk without any fixes
ci_basic3 <- apply(sim_data3, 1, bootstrap_ci, method = "basic", B = 1000, summary_fn = mean, na.rm = TRUE)
# don't need any ... arguments here, but illustrating the idea of the ...
# notice that bootstrap_ci has default seed = 100 and C = 0.95 arguments
ci_basic_df3 <- as.data.frame(t(ci_basic3))
names(ci_basic_df3) <- c("lower", "upper")
ci_basic_df3 %>% mutate(covered = lower <= pop_mean3 & upper >= pop_mean3) %>%
  summarize(coverage_probability = mean(covered))
```

```{r boot normal CI 3}
# You should be able to just run this code chunk without any fixes
ci_norm3 <- apply(sim_data3, 1, bootstrap_ci, method = "normal", B = 1000, summary_fn = mean, na.rm = TRUE)
# don't need any ... arguments here, but illustrating the idea of the ...
# notice that bootstrap_ci has default seed = 100 and C = 0.95 arguments
ci_norm_df3 <- as.data.frame(t(ci_norm3))
names(ci_norm_df3) <- c("lower", "upper")
ci_norm_df3 %>% mutate(covered = lower <= pop_mean3 & upper >= pop_mean3) %>%
  summarize(coverage_probability = mean(covered))
```

## Simulation Part 4 - Assumptions Not Met, Small Samples

Simulate 1000 sets of 10 random numbers from a $Exp(1)$ distribution and store them in a 1000 x 100 matrix `sim_data4`. Note that `sim_data4` represents 1000 samples of size 10 under conditions where we *know* the assumptions of a t confidence interval are not met and Central Limit Theorem is almost certainly *not* going to cover us.

Copy and modify your code chunks from Simulation Part 1 to produce 95% confidence intervals using `sim_data4`. What proportion of "95% confidence intervals" actually contained the true parameter value? Did any methods perform noticeably better/worse?

```{r sim 4}
set.seed(437)
# need to do the simulation now!
sim_data4 <- matrix(rexp(1000*10), nrow = 1000, ncol = 10)
```

```{r t-CI4}
# You should be able to just run this code chunk without any fixes
pop_mean4 <- 1
ci_t4 <- apply(sim_data4, 1, function(x) t.test(x)$conf.int)
ci_t_df4 <- as.data.frame(t(ci_t4))
names(ci_t_df4) <- c("lower", "upper")
ci_t_df4 %>% mutate(covered = lower <= pop_mean4 & upper >= pop_mean4) %>%
  summarize(coverage_probability = mean(covered))
```

```{r boot percentile CI 4}
# You should be able to just run this code chunk without any fixes
ci_perc4 <- apply(sim_data4, 1, bootstrap_ci, method = "percentile", B = 1000, summary_fn = mean, na.rm = TRUE)
# don't need any ... arguments here, but illustrating the idea of the ...
# notice that bootstrap_ci has default seed = 100 and C = 0.95 arguments
ci_perc_df4 <- as.data.frame(t(ci_perc4))
names(ci_perc_df4) <- c("lower", "upper")
ci_perc_df4 %>% mutate(covered = lower <= pop_mean4 & upper >= pop_mean4) %>%
  summarize(coverage_probability = mean(covered))
```

Copy and modify the chunk above for the basic and normal-theory intervals.

```{r boot basic CI 4}
# You should be able to just run this code chunk without any fixes
ci_basic4 <- apply(sim_data4, 1, bootstrap_ci, method = "basic", B = 1000, summary_fn = mean, na.rm = TRUE)
# don't need any ... arguments here, but illustrating the idea of the ...
# notice that bootstrap_ci has default seed = 100 and C = 0.95 arguments
ci_basic_df4 <- as.data.frame(t(ci_basic4))
names(ci_basic_df4) <- c("lower", "upper")
ci_basic_df4 %>% mutate(covered = lower <= pop_mean4 & upper >= pop_mean4) %>%
  summarize(coverage_probability = mean(covered))
```

```{r boot normal CI 4}
# You should be able to just run this code chunk without any fixes
ci_norm4 <- apply(sim_data4, 1, bootstrap_ci, method = "normal", B = 1000, summary_fn = mean, na.rm = TRUE)
# don't need any ... arguments here, but illustrating the idea of the ...
# notice that bootstrap_ci has default seed = 100 and C = 0.95 arguments
ci_norm_df4 <- as.data.frame(t(ci_norm4))
names(ci_norm_df4) <- c("lower", "upper")
ci_norm_df4 %>% mutate(covered = lower <= pop_mean4 & upper >= pop_mean4) %>%
  summarize(coverage_probability = mean(covered))
```

## Simulation Part 5 - Range Preservation

Produce a histogram of the lower bound of the CIs you obtained using each of the four methods in Simulation Part 4. Which methods appear to be range-preserving (that is, cannot give implausible values for the parameter) even under this "worst-case scenario"? 

```{r hist }
hist(x = ci_t_df4$lower)
hist(ci_perc_df4$lower)
hist(ci_basic_df4$lower)
hist(ci_norm_df4$lower)
```
The only method that is range preserving is percentile because the others give p values less than 0.


